Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2017-06-07 paper/paper.tex
made some figures pdf

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, room 1005, New York, NY 10003, USA}
\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}

\author[0000-0002-3962-9274]{Boris Leistedt}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, room 1005, New York, NY 10003, USA}
\affil{NASA Einstein Fellow}

@@

\section{Data and results}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{data.png}
\caption{The observed \cmd: To visualize the full dataset, in the left panel we use the point estimates of the \tmass\ $J-K_s$ color, the $J$ band apparent magnitude, and the \gaia\ parallax. The blue lines represent the $1\sigma$ and $2\sigma$ contours for the distribution. To give a sense of the uncertainties, in the right panel we subsample the dataset and include the $1\sigma$ uncertainties. The uncertainties are dominated by the parallax noise, with many stars diverging to infinitely far away and therefore very intrinsically bright.}
\label{fig:data}
\end{figure}

@@

\subsection{The empirical prior}
Using \xd\ and the MMLE method described in the methods section, we generate the
prior, shown in \figurename~\ref{fig:prior}.
The left panel shows 1.3 million samples from the prior distribution, and the
right panel shows $1\sigma$ contours for each component in the underlying
Gaussian mixture model.
Compared to the raw data in \figurename~\ref{fig:data}, the deconvolved
color--magnitude diagram for the \gaia\ + \tmass\ stars is tighter, as expected.
We'll come back to more features in the data below, especially those seen in the posterior distribution.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{prior_ngauss128.png}
\caption{The \cmd\ prior, modeled as a Gaussian mixture inferred by
  running XD on all of the data. The left panel shows a sampling of
  the prior, with the blue lines showing the $1\sigma$ and $2\sigma$
  contours. The right panel shows the 1-$\sigma$ contours of the
  individual components. The latter are Gaussian in the transformed
  magnitude space to make the XD inference tractable. Thus, they
  appear as slightly deformed ellipses in color--magnitude space.}
\label{fig:prior}
\end{figure}

@@

This change in precision is further illustrated in Figure~\ref{fig:delta}, where we show the fractional changes in the variance. The left panel shows the natural log of the fractional change in variance as a function of $(J-K_s)^C$ color, with the $1\sigma$ and $2\sigma$ contours over plotted.
The points are colored by the natural log of their fractional change in variance (the y-axis of the left panel) in both panels to help guide your eye for the right panel. The regions of \cmd\ space with the greatest improvements are shown in black, and the regions with the lowest improvements are shown in yellow.
The giants show obvious large improvements, as well as some main sequence stars.
Regions of large improvement are those with the largest photometric and parallax errors, which are the faintest observed objects.
This is a fairly standard result which we illustrated in the toy model above: the deconvolution of uncertainties is stronger in noisier regions of the data, leading to narrower features in the upper part of the \cmd, for example.
Stars with a color $(J-K_s)^C \sim 0.6$ show a decrease in precision due to the width of the prior there; the more vertical structure in the red giant branch of the \cmd.

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.png}
\caption{Five example posterior PDFs over parallax using our method.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented by a mixture of Gaussians in green.
We chose example points that evenly spanned the color space, but within each color bin, we chose a point at random.
The remaining panels show the posterior PDF in blue, the prior PDF in green, and likelihood of the data in black.
Most stars have posterior PDFs (blue) that are only slightly more precise than the data (black), which is well represented in this random sampling of points.
Examples 1-4 are not very effected by the prior (green), where good data is good data.
Example 4 is interesting because it lies in the same color range as the red clump, where there is a lot of support from the prior, but the likelihood does not have much support there so the parallax posterior PDF, transformed into an absolute magnitude, remains on the main sequence.
Example 5 shows a significantly more precise posterior PDF compared with the likelihood, and looks very similar to the prior PDF.
It also exemplifies that our prior tends to have the largest effect on stars in the red giant branch.}
\label{fig:posterior}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posteriorCMD.png}
\caption{The distribution of expectation values of the parallax posterior PDFs transformed into an absolute magnitude vs color, the \cmd, of the full data set is shown in the left panel, with the $1\sigma$ and $2\sigma$ contours of the distribution shown in black.
The right panel shows a subsample of data set, plotting the expectation value and $1\sigma$ uncertainty of posterior PDF projected onto the \cmd.
Notice that some parts of this \cmd\ are tighter than the prior, or noise-deconvolved \cmd, shown in \figurename~\ref{fig:prior}
The visual of the prior is a sampling of the Gaussians, but this is the expectation
  value, so it lies much closer to the center of the
  distribution. Similar to the toy model, where the maximum of the
  posterior PDF of the noisy points lies closer to the center of the true distribution than the true values.}
\label{fig:posteriorCMD}
\end{figure}

@@

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.
M67 is at a Heliocentric distance of $800$--$900~{\rm pc}$ with an estimated age of $3$--$5~{\rm Gyr}$.
Previous work \citep{cbj15} has noted that distance estimates directly derived from the inverse of the observed (noisy) parallax are unreliable (significantly biased and uncertain) when the parallax signal-to-noise (SN) is below 5, $\varpi / \sigma_\varpi < 5$.
With \gaia\ parallaxes having a minimum uncertainty of about $0.3~{\rm mas}$, the SN reaches 5 for the most certain parallax measurements at about $700~{\rm pc}$, just before the distance to this cluster.
M67 is therefore a great test case for comparing parallax inferences with or without a prior.

@@

\subsection{Comparison with other priors}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{comparePrior.png}
\caption{Comparison with a common spatial prior: A visualization of the posterior PDF for a subsection of \tgas\ stars, shown as the expectation value and the sqrt of the variance as the uncertainty. Left panel shows the posterior PDF using our \cmd\ prior, right panel shows the posterior PDFs when using the exponentially declining spatial density, a common space-density prior used when inferring distances. The scale of the exponentially declining spatial density prior is $1.35~{\rm kpc}$, the optimal value found in \cite{astraatmadja16a}. Using the \cmd\ as a prior generates a posterior PDF that has smaller variance.}
\label{fig:comparePrior}
\end{figure}

>>>>

\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Computational Astrophysics, Flatiron Institute, 162 Fifth Ave, New York, NY 10010, USA\hspace*{-1cm}}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
\affil{Center for Data Science, New York University, 60 Fifth Ave, New York, NY 10011, USA}
\affil{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg}

\author[0000-0002-3962-9274]{Boris Leistedt}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University, 726 Broadway, New York, NY 10003, USA}
\affil{NASA Einstein Fellow}

@@

\section{Data and results}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{data.pdf}
\caption{The observed \cmd: To visualize the full dataset, in the left panel we use the point estimates of the \tmass\ $J-K_s$ color, the $J$ band apparent magnitude, and the \gaia\ parallax. The blue lines represent the $1\sigma$ and $2\sigma$ contours for the distribution. To give a sense of the uncertainties, in the right panel we subsample the dataset and include the $1\sigma$ uncertainties. The uncertainties are dominated by the parallax noise, with many stars diverging to infinitely far away and therefore very intrinsically bright.}
\label{fig:data}
\end{figure}

@@

\subsection{The empirical prior}
Using \xd\ and the MMLE method described in the methods section, we generate the
prior, shown in \figurename~\ref{fig:prior}.
The left panel shows 1.3 million samples from the prior distribution, and the
right panel shows $1\sigma$ contours for each component in the underlying
Gaussian mixture model.
Compared to the raw data in \figurename~\ref{fig:data}, the deconvolved
color--magnitude diagram for the \gaia\ + \tmass\ stars is tighter, as expected.
We'll come back to more features in the data below, especially those seen in the posterior distribution.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{prior.pdf}
\caption{The \cmd\ prior, modeled as a Gaussian mixture inferred by
  running XD on all of the data. The left panel shows a sampling of
  the prior, with the blue lines showing the $1\sigma$ and $2\sigma$
  contours. The right panel shows the 1-$\sigma$ contours of the
  individual components. The latter are Gaussian in the transformed
  magnitude space to make the XD inference tractable. Thus, they
  appear as slightly deformed ellipses in color--magnitude space.}
\label{fig:prior}
\end{figure}

@@

This change in precision is further illustrated in Figure~\ref{fig:delta}, where we show the fractional changes in the variance. The left panel shows the natural log of the fractional change in variance as a function of $(J-K_s)^C$ color, with the $1\sigma$ and $2\sigma$ contours over plotted.
The points are colored by the natural log of their fractional change in variance (the $y$ axis of the left panel) in both panels to help guide your eye for the right panel. The regions of \cmd\ space with the greatest improvements are shown in black, and the regions with the lowest improvements are shown in yellow.
The giants show obvious large improvements, as well as some main sequence stars.
Regions of large improvement are those with the largest photometric and parallax errors, which are the faintest observed objects.
This is a fairly standard result which we illustrated in the toy model above: the deconvolution of uncertainties is stronger in noisier regions of the data, leading to narrower features in the upper part of the \cmd, for example.
Stars with a color $(J-K_s)^C \sim 0.6$ show a decrease in precision due to the width of the prior there; the more vertical structure in the red giant branch of the \cmd.

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.png}
\caption{Five example posterior PDFs over parallax using our method.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented by a mixture of Gaussians in green.
We chose example points that evenly spanned the color space, but within each color bin, we chose a point at random.
The remaining panels show the posterior PDF in blue, the prior PDF in green, and likelihood of the data in black.
Most stars have posterior PDFs (blue) that are only slightly more precise than the data (black), which is well represented in this random sampling of points.
Examples 1 through 4 are not very effected by the prior (green), where good data is good data.
Example 4 is interesting because it lies in the same color range as the red clump, where there is a lot of support from the prior, but the likelihood does not have much support there so the parallax posterior PDF, transformed into an absolute magnitude, remains on the main sequence.
Example 5 shows a significantly more precise posterior PDF compared with the likelihood, and looks very similar to the prior PDF.
It also exemplifies that our prior tends to have the largest effect on stars in the red giant branch.}
\label{fig:posterior}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posteriorCMD.pdf}
\caption{The distribution of expectation values of the parallax posterior PDFs transformed into an absolute magnitude vs color, the \cmd, of the full data set is shown in the left panel, with the $1\sigma$ and $2\sigma$ contours of the distribution shown in black.
The right panel shows a subsample of data set, plotting the expectation value and $1\sigma$ uncertainty of posterior PDF projected onto the \cmd.
Notice that some parts of this \cmd\ are tighter than the prior, or noise-deconvolved \cmd, shown in \figurename~\ref{fig:prior}
The visual of the prior is a sampling of the Gaussians, but this is the expectation
  value, so it lies much closer to the center of the
  distribution. Similar to the toy model, where the maximum of the
  posterior PDF of the noisy points lies closer to the center of the true distribution than the true values.}
\label{fig:posteriorCMD}
\end{figure}

@@

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.

M67 is at a Heliocentric distance of $800$--$900~{\rm pc}$ with an estimated age of $3$--$5~{\rm Gyr}$.
Previous work \citep{cbj15} has noted that distance estimates directly derived from the inverse of the observed (noisy) parallax are unreliable (significantly biased and uncertain) when the parallax signal-to-noise (SN) is below 5, $\varpi / \sigma_\varpi < 5$.

With \gaia\ parallaxes having a minimum uncertainty of about $0.3~{\rm mas}$, the SN reaches 5 for the most certain parallax measurements at about $700~{\rm pc}$, just before the distance to this cluster.
M67 is therefore a great test case for comparing parallax inferences with or without a prior.

@@

\subsection{Comparison with other priors}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{comparePrior.pdf}
\caption{Comparison with a common spatial prior: A visualization of the posterior PDF for a subsection of \tgas\ stars, shown as the expectation value and the sqrt of the variance as the uncertainty. Left panel shows the posterior PDF using our \cmd\ prior, right panel shows the posterior PDFs when using the exponentially declining spatial density, a common space-density prior used when inferring distances. The scale of the exponentially declining spatial density prior is $1.35~{\rm kpc}$, the optimal value found in \cite{astraatmadja16a}. Using the \cmd\ as a prior generates a posterior PDF that has smaller variance.}
\label{fig:comparePrior}
\end{figure}
